---
title: "CS625-HW5"
author: "Syamala Akhila Devi Bolla"
date: "2023-03-22"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part 1: Create Distribution Charts


* I choosed the first dataset - [US Census Bureau County Population dataset](https://www.census.gov/data/datasets/time-series/demo/popest/2010s-counties-total.html#par_textimage_70769902) (CO-EST2019-alldata) which has 3194 rows and 164 columns. 

#### Question1: For the boxplot, show the distribution of the population of all counties in the US as of the 2010 Census (`CENSUS2010POP`), the 2015 estimate (`POPESTIMATE2015`), and the 2019 estimate (`POPESTIMATE2019`) (should result in 3 separate boxplot glyphs in a single chart)

### Boxplots (box-and-whisker diagrams):

* Boxplots are useful and convenient for summarizing and visualizing the distribution of a dataset, and can help to identify patterns, trends, and outliers that may not be immediately apparent from the raw data. Box Plots can be drawn either vertically or horizontally.

### Importing the Data

```{r}
data <- read.csv(file="./co-est2019-alldata.csv")
df <- subset(data)
```

* The data is imported into data variable that displays data in different format. So, format the data using subset. Then, we will get data in proper rows and columns. 

* Converted the data and stored in df and the data is displayed which has 164 columns.

* To Perform the given task we need to filter out only 2010 Census (CENSUS2010POP), 2015 estimate (POPESTIMATE2015), 2019 estimate (POPESTIMATE2019) columns.

```{r}
df_1 <- df[, c("CENSUS2010POP", "POPESTIMATE2015", "POPESTIMATE2019")];
head(df_1)
```

* The data is filtered and displayed the contents of data.

* Using R functions plot the box plot for the data, added title as “POPULATION CENSUS OF 2010 WITH THE POPULATION ESTIMATION OF 2015 and 2019” and the x axis element labels as “2010 CENSUS”, “2015 ESTIMATE”, “2019 ESTIMATE” which comes under "YEAR" and y axis labels as "CENSUS VALUE".

* The plot is shown below:

```{r}
label = c("2010 CENSUS", "2015 ESTIMATE", "2019 ESTIMATE");
boxplot(df_1, main="POPULATION CENSUS OF 2010 WITH THE POPULATION ESTIMATION OF 2015 and 2019", ylab="CENSUS VALUE", xlab="YEAR",names=label)
```
### Observation:

* The output is not as expected, the three glyph’s in the box plot are not clear enought too see five points.

* There are significant outliers on the upper side of the glyph.

* According to the graph, all the five points (min, 25%, median, 75%, and max) are marked on one single line.

* We have a option to plot the data without the outliers which helps us to look at the data and get the five point values.

### Advantages:

* No Advantages since the Visualization of the glyph is poor.

### Disadvantages:

* I could not identify all the statistical values using the boxplot.

* The plot need a manipulation for better visualization.

* The range and IQR could not be distinguished.


```{r}
label = c("2010 CENSUS", "2015 ESTIMATE", "2019 ESTIMATE")
boxplot(df_1,outline=FALSE, main="POPULATION CENSUS OF 2010 WITH THE POPULATION ESTIMATION OF 2015 and 2019", ylab="CENSUS VALUE", xlab="YEAR", names=label)
```

### Observations:

* From the above boxplot, we could see that the outliers have been removed and all the five points(min, 25%, median, 75%, and max) can differentiated.

* The population count is very clear, and I could tell the approximate value at each of the five required points.

### Advantages:

* By eliminating the outliers, five statistical points have been identified.

* The Inter Quartile Range (IQR) is clearly identified.

* The values can be easily compared between three glyphs.

* Median values can also be identified.


### Disadvantages:

* We don’t know how accurate is the output because the outliers had huge impact on previous graph.


#### Question2: For the eCDF and histogram, show the distribution of the population of all counties in the US as of the 2010 Census (`CENSUS2010POP` column)
   
### eCDF (Empirical Cumulative Distribution Function):

* An empirical cumulative distribution function (eCDF) is a non-parametric estimator of the cumulative distribution function (CDF) of a random variable. It is a step function that gives the proportion or fraction of data points that are less than or equal to a certain value. To construct an eCDF, the data points are first sorted in ascending order.

* Only the 2010 Census population (CENSUS2010POP) is assigned to df_2 variable.

```{r}
 df_2 <- df[ , c("CENSUS2010POP")]
```

* Imported the reshape library for using melt() function.

```{r}
library(reshape2)
```

* The melt() function in R is part of the reshape2 package and it allows you to reshape a wide-format data frame into a long-format data frame. The melt() function takes a wide-format data frame and reorganizes it into a long-format data frame.

```{r}
data_long <- melt(df_2)               #Reshaping data frame
head(data_long)                       #Print head of long data
```

```{r}
ecdf(df_2)
```

```{r}
plot(ecdf(df_2),main="ECDF of 2010 POPULATION CENSUS")
```

### Observation:

* X-axis and Y axis are quantitative values with CENSUS2010POP and Percentage.

* I could see a continuous step graph for the values across the x-axis and percentages across the y-axis.

* Most of the data points are concentrated within a small range and the data is scattered from two-digit number to millions, it appears to have less steps.

### Histogram:

* The number of bins in a histogram can be chosen independently of the number of items in the dataset. The purpose of a histogram is to display the distribution of a continuous variable, so the data is divided into a specified number of equally spaced bins along the x-axis, and the height of each bar represents the number of observations that fall within each bin. Choosing the appropriate number of bins for a histogram is important because it can affect how the data is perceived. 

```{r}
hist(df_2,main="DISTRIBUTION OF POPULATION IN 2010",breaks=10)
```

### Observations:

* The view of bins, x-axis, and y-axis are not as expected.

* The title of graph has been set to “DISTRIBUTION OF POPULATION IN 2010”.

* The bins breaks are 10 as given in the condition.

* The above histogram looks right skewed and most of the values seem concentrated in only one bin.

* Rest of the data points could not be clearly identified on the histogram.

* I could tell the approximate number by looking at the first bin and the y-axis, but I could not tell for the rest of the bins.

* The x-axis is scaled 0-millions.

* The y-axis has a number count from 0-2500.

* We knew there are 3193 counties with different population size. However, in this graph, the output is not clear because of the outliers.

```{r}
hist(df_2,main="DISTRIBUTION OF POPULATION IN 2010",breaks=100)
```

### Observation:

* The histogram is still right skewed and most of the values seems to be in only one bin.

* Rest of the data points could not be clearly identified on the histogram.

* The x-axis is scaled 0-millions.

* The y-axis has a number count from 0-3000.

* Maximum counties are still concentrated in one bin.


## Part 2: Further Analysis

### Histogram

* The above histogram with different bin sizes is not clearly showing the data distribution. Let's try removing the outliers which may give us the clear picture of data.
* Let's write a trimming function that removes outliers.

```{r}
trim_h <- function(x){x[(x>quantile(x, 0.25) - 1.5 * IQR(x)) & (x<quantile(x, 0.75) + 1.5 * IQR(x))]}

hist(trim_h(df_2), main="Histogram of 2010 Population Census")
```

### Observations:

* After removing the outliers, we can easily see the distribution of population over several counties.
* The bins, x-axis, and y-axis are displayed as expected.

### Advantages:

* Now, Able to compare data easily.
* Able to differentiate range of data in each bin.
* Able to easily find the approximate number of people with corresponding values.
* Able to easily identify the highest and lowest values.

### Disadvantages:

* It's difficult to tell the exact value corresponding to each bin.

### More Examples

* The histogram with smaller bin size gives clear visualization of the data distribution.

```{r}
hist(trim_h(df_2),main="Histogram of 2010 Population Census",breaks=50)
```

#### ECDF

* Used the same trimming function to see whether any difference in ECDF.

```{r}
plot(ecdf(trim_h(df_2)))                                  #Create ecdf plot in R
```

### Observations:

* Able to fit the whole data in the definite axis.

* By modifying the data size and by eliminating the outliers, the data distribution is better.

* Corresponding value of each distribution could not be identified.

* The continuity between the range of each interval could not be identified.

### Conclusion

* The data which I worked on had a lot of outliers that made the evaluation of data distribution hard.

* After removing outliers, among the three different idioms (Boxplots, ECDF, Histogram), the histogram with a smaller number of bins gave a clear visualization of the data distribution.


## References

* Reference 1: https://datavizcatalogue.com/methods/histogram.html

* Reference 2: https://statisticsglobe.com/draw-multiple-boxplots-in-one-graph-in-r

* Reference 3: https://stackoverflow.com/questions/53833773/how-to-make-a-trimmed-histogram-in-r

* Reference 4: https://datavizcatalogue.com/methods/box_plot.html

* Reference 5: https://stackoverflow.com/questions/14207739/how-to-remove-outliers-in-boxplot-in-r

* Reference 6: https://statisticsglobe.com/extract-certain-columns-of-data-frame-in-r

* Reference 7: https://edwinth.github.io/blog/outlier-bin/

* Reference 8: https://stackoverflow.com/questions/53833773/how-to-make-a-trimmed-histogram-in-r



